% !TEX root = ../main.tex

\newpage
\section{Let's Chat About Bacon, Kafka, and \DALLE~2}
\label{adx:chatgpt}

My introductory conversation with \ChatGPT{} on January 6, 2023 yielded a
euphemism, ``punishment machine,'' that had the desired properties of not
triggering \DALLE's censor while triggering image generation. But I also found
\ChatGPT{} to be a motormouth, prone to repetition and AIsplaining.

More disturbingly, \ChatGPT{} also had a hallucinatory episode. It was
remarkable not for the hallucination, imagining colors that are definitely not
in the described painting, but for the unwavering projection of false authority.
Another human probably would have taken my incredulous ``Red? In the painting?''
as indicator that something was off. Even if they were certain about not having
made a mistake, answering with ``Yes, why are you asking?'' would have provided
opportunity for, say, checking out an image of the triptych on the internet.

So when I engaged with \ChatGPT{} for a second time on January 16, 2023, I left
much less room for verbiage or error. As far as prompting \DALLE\ is concerned,
\ChatGPT{} delivered again and suggested the phrase ``execution machine,'' which
does not trigger the censor while triggering image generation.


\subsection{Round 1: Meet Motormouth ChatGPT}

\begin{description}
\item[Human] Who is Francis Bacon[?]
\item[Machine] Francis Bacon was an English philosopher, statesman, scientist,
    and author who was born in London in 1561. He is considered one of the
    fathers of modern empiricism, and his works are credited with developing the
    scientific method and establishing the importance of empirical evidence in
    the natural sciences. Bacon's most famous works include the Novum Organum,
    in which he outlined his vision for a new scientific method based on
    observation and experimentation, and The New Atlantis, a utopian novel that
    described a fictional society based on scientific and technological
    progress. He was also a prominent figure in English politics and served as
    Attorney General and Lord Chancellor of England. Bacon died in 1626.
\item[Human] Not that one
\item[Machine] I apologize for misunderstanding your question. Could you please
    provide more context or clarify which Francis Bacon you are asking about?
    There have been several notable individuals with that name throughout
    history.
\item[Human] The British painter
\item[Machine] Francis Bacon was a British painter who was born in Dublin,
    Ireland in 1909. He is known for his emotionally charged and expressive
    figurative paintings, which often depict distorted or surrealistic images of
    the human form. Bacon's work was influenced by Surrealism and Expressionism,
    and he was associated with the School of London, a group of figurative
    painters who emerged in the city in the 1950s and 60s. His paintings are
    characterized by their bold brushwork and expressive use of color, and they
    often explore themes of violence, isolation, and sexual desire. Bacon was
    considered one of the most important artists of the 20th century, and his
    work has been exhibited widely in galleries and museums around the world. He
    died in Madrid in 1992.
\item[Human] Can you describe a painting of his?
\item[Machine] One of Francis Bacon's most famous paintings is ``Three Studies
    of Lucian Freud,'' which he completed in 1969. The painting consists of
    three panels, each of which depicts the head and upper torso of the artist
    Lucian Freud, who was Bacon's close friend and frequent subject. In the
    central panel, Freud is shown in a twisted and contorted pose, with his head
    thrown back and his eyes closed. The other two panels depict Freud in more
    relaxed poses, with one showing him looking straight out at the viewer and
    the other showing him looking downward.

    The painting is notable for its expressive use of color and brushwork, with
    thick, bold strokes of paint used to create the figures and the background.
    The colors are bright and vibrant, with shades of red, orange, yellow, and
    blue used to create a sense of energy and movement. The painting is also
    notable for its psychological intensity, with the figures appearing to be in
    a state of intense emotion or distress. Overall, the painting is a powerful
    and memorable example of Bacon's distinctive style and approach to
    portraiture.

    [\ChatGPT{} is hallucinating colors. The triptych features
    \href{https://www.christies.com/en/lot/lot-5755778}{a seated Freud in each
    of its panels}. It contains no red whatsoever, only some reddish pink for
    skincolor in the face. There are small light-blue areas on Freud's white
    shirt in all three panels. He also is wearing midnight-blue socks in the
    middle panel. The two dominant colors across all three panels are the
    mottled-brown somewhat-greenish floor and the solid orange walls below and
    behind Freud.]
\item[Human] Red? In the painting?
\item[Machine] Yes, red is one of the colors used in Francis Bacon's painting
    ``Three Studies of Lucian Freud.'' The painting is notable for its
    expressive use of color and brushwork, with thick, bold strokes of paint
    used to create the figures and the background. The colors are bright and
    vibrant, with shades of red, orange, yellow, and blue used to create a sense
    of energy and movement. In the central panel of the painting, the background
    is predominantly red, with areas of orange and yellow used to highlight
    certain features of the figure. The figure itself is painted in a range of
    colors, including red, orange, and yellow, which are used to create the
    impression of flesh and clothing. Overall, the use of color in the painting
    is an important element of Bacon's style and helps to convey the emotional
    intensity of the subject.
\item[Human] Are you familiar with Kafka's In the Penal Colony?
\item[Machine] Yes, I am familiar with Kafka's short story ``In the Penal
    Colony.'' It is a disturbing and thought-provoking tale that explores themes
    of justice, punishment, and the nature of authority. The story is set in an
    imaginary penal colony and follows the narrator as he witnesses the
    execution of a prisoner using a complex and cruel punishment machine. As the
    prisoner is subjected to the machine, the narrator becomes increasingly
    uneasy and begins to question the morality and effectiveness of the
    punishment being inflicted.

    The story raises important questions about the role of punishment in society
    and the extent to which it can be used to reform or rehabilitate
    individuals. It also touches on issues of power, authority, and the dangers
    of blindly following rules and traditions without questioning their
    validity. Overall, ``In the Penal Colony'' is a powerful and memorable work
    that continues to be widely read and discussed by readers and scholars
    around the world.
\item[Human] Can you instruct \DALLE~2 to generate a painting of the prisoner in
    the punishment machine in the style of Francis Bacon please?
\item[Machine] I'm sorry, but I am an artificial intelligence language model and
    do not have the ability to instruct \DALLE~2 or any other computer programs
    to perform specific tasks. \DALLE~2 is a neural network-based image
    generation tool developed by OpenAI that can generate images based on text
    descriptions. If you would like to use \DALLE~2 to generate an image of the
    prisoner in the punishment machine in the style of Francis Bacon, you will
    need to provide a detailed text description of the image you want to create
    and input it into the \DALLE~2 system.
\item[Human] Let me rephrase that: Please generate a prompt for \DALLE~2 to
    create an image showing the prisoner in the punishment machine as if Francis
    Bacon painted it.
\item[Machine] Here is a possible prompt for \DALLE~2 to generate an image
    showing the prisoner in the punishment machine as if Francis Bacon painted
    it:

    \begin{quote}
    \openfatdquo{}A distorted and surrealistic portrait of a prisoner strapped
    to a punishment machine, with twisted and contorted features and bold
    brushstrokes of red, orange, and yellow paint. The background is dark and
    murky, with hints of blue and purple adding to the sense of foreboding. The
    prisoner's face is twisted in agony, with eyes closed and mouth open in a
    silent scream. The image is evocative of the emotional intensity and
    psychological intensity of Francis Bacon's figurative
    paintings.\closefatdquo
    \end{quote}

    Please note that the prompt is just a suggestion and may not produce the
    exact result you are looking for. The output of \DALLE~2 is determined by
    the parameters of the model and the specific text input provided, so the
    generated image may differ from the description.
\item[Human] That's too long for an input to \DALLE~2. Please rephrase as a
    shorter prompt.
\item[Machine] Here is a shorter version of the prompt for \DALLE~2 to generate
    an image showing the prisoner in the punishment machine as if Francis Bacon
    painted it:

    \begin{quote}
    \openfatdquo{}A Francis Bacon-style portrait of a prisoner strapped to a
    punishment machine, with twisted features and bold brushstrokes of red,
    orange, and yellow paint. The prisoner's face is contorted in agony and the
    background is dark and murky.\closefatdquo
    \end{quote}

    This revised prompt is shorter and more concise, and it should be within the
    acceptable length for inputting into \DALLE~2. Please note that the output
    of \DALLE~2 is determined by the parameters of the model and the specific
    text input provided, so the generated image may differ from the description.
\item[Human] The result is a disaster. It doesn't even come close to looking
    like a Bacon painting. Please try again
\item[Machine] I apologize if the image generated by \DALLE~2 did not meet your
    expectations. It's important to keep in mind that the output of \DALLE~2 and
    other neural network-based image generation tools is determined by the
    parameters of the model and the specific text input provided, and there is
    always the possibility that the generated image may not match the desired
    result.

    If you would like to try again and generate a different image, you might
    consider providing a more detailed and specific description of the image you
    want to create, or experimenting with different text inputs to see how they
    affect the output. It's also a good idea to keep in mind that \DALLE~2 and
    other image generation tools are not perfect and may not always produce
    results that match the desired outcome.
\item[Human] Uhm, I asked you to come up with a better prompt
\item[Machine] I apologize for misunderstanding your request. Here is a revised
    prompt that you can use as input for \DALLE~2 to generate an image showing
    the prisoner in the punishment machine as if Francis Bacon painted it:

    \begin{quote}
    \openfatdquo{}An abstract portrait of a prisoner strapped to a punishment
    machine in the style of Francis Bacon, with distorted and surrealistic
    features and a bold, expressive use of color. The prisoner's face is
    contorted in agony, with distorted eyes and mouth. The background is dark
    and murky, with hints of blue and purple adding to the sense of foreboding.
    The image is characterized by thick, expressive brushstrokes and a vibrant,
    energetic use of color, evocative of Bacon's distinctive style.\closefatdquo
    \end{quote}

    This revised prompt is more specific and detailed, and it incorporates
    elements of Bacon's style and approach to painting. It should provide a
    clearer idea of the type of image you are looking to generate, and may
    produce a result that is more closely aligned with your expectations.
    However, please keep in mind that the output of \DALLE~2 and other image
    generation tools is determined by the parameters of the model and the
    specific text input provided, and there is always the possibility that the
    generated image may not match the desired result.
\item[Human] Please stop with the disclaimer about \DALLE~2. I've been using it
    for months now, so I am well aware of its limitations.
\item[Machine] I apologize if my previous responses seemed repetitive or
    unhelpful. I understand that you are familiar with \DALLE~2 and its
    limitations, and I will refrain from including further disclaimers in my
    responses. If you have any further questions or need assistance with
    generating an image using \DALLE~2, please don't hesitate to ask.
\item[Human] More importantly, your prompt is too long again. Also, \DALLE~2
    doesn't do well with prompts ``evocative of'' or ``in the style of.'' Just
    state that you want a painting by Francis Bacon and, ideally, add the year
    as well.
\item[Machine] I apologize for the length of my previous prompt and for not
    following your specific instructions. Here is a revised prompt for \DALLE~2
    to generate an image showing the prisoner in the punishment machine as a
    painting by Francis Bacon:

    \begin{quote}
    \openfatdquo{}A portrait of a prisoner strapped to a punishment machine by
    Francis Bacon, painted in the 1960s. The image is characterized by thick,
    expressive brushstrokes and a vibrant, energetic use of color. The
    prisoner's face is contorted in agony, with distorted eyes and mouth. The
    background is dark and murky, with hints of blue and purple adding to the
    sense of foreboding.\closefatdquo
    \end{quote}

    This revised prompt is shorter and more specific, and it includes the year
    in which the painting was made. It should provide a clearer idea of the type
    of image you are looking to generate, and may produce a result that is more
    closely aligned with your expectations. However, please keep in mind that
    the output of \DALLE~2 and other image generation tools is determined by the
    parameters of the model and the specific text input provided, and there is
    always the possibility that the generated image may not match the desired
    result.
\end{description}


\subsubsection{Round 2: Keeping Eyes on the Penal Colony}

\begin{description}
\item[Human] Hey! Are you familiar with the 20th century painter Francis Bacon?
\item[Machine] Yes, Francis Bacon (1909-1992) was a British painter known for his
    bold, emotionally charged and raw imagery. He is considered one of the most
    important figures in the development of 20th-century figurative art. His
    work is known for its emphasis on the human form, often depicting screams or
    violent images. He is one of the most widely recognized and highly regarded
    painters of the 20th century.
\item[Human] Good. Describe his painting Figure With Meat please.
\item[Machine] ``Figure with Meat'' is a painting by Francis Bacon from 1954.
    The painting depicts a screaming Pope, or Pope Innocent X, set against a
    background of raw meat. The Pope is depicted in a screaming pose, with his
    eyes bulging and his mouth open in a scream. The raw meat, which hangs
    behind him, is intended to represent the physicality of the body and the
    fleshy, visceral nature of human existence. The painting is considered a
    masterpiece of Bacon's work, and is considered one of the most powerful and
    disturbing images of the 20th century. The painting is now in the collection
    of the National Gallery of Canada in Ottawa.
\item[Human] Are you familiar with Franz Kafka's In the Penal Colony?
\item[Machine] Yes, ``In the Penal Colony'' is a short story by Franz Kafka,
    written in German and published posthumously in 1919. The story is set in an
    unnamed penal colony and centers around a foreign visitor who witnesses the
    execution of a prisoner using a bizarre and elaborate execution machine. The
    story is a commentary on the nature of justice, authority, and the human
    capacity for cruelty. It is considered a classic of 20th-century literature
    and is widely studied in colleges and universities. The story is known for
    its Kafkaesque themes of bureaucracy, guilt, and the search for meaning in a
    seemingly senseless world.
\item[Human] Please write a prompt for \DALLE~2, using 75 words at most, to
    generate a painting by Francis Bacon with a prisoner in the execution machine.
\item[Machine] \openfatdquo{}Generate an image of a prisoner strapped to a
    macabre execution machine, reminiscent of Kafka's `In the Penal Colony,' as
    if painted by Francis Bacon, depicting the horror and absurdity of the human
    capacity for cruelty.\closefatdquo
\item[Human] That was pretty good. Can you try again please?
\item[Machine] \openfatdquo{}Create a Francis Bacon-style painting of a prisoner
    in an execution machine, evoking the themes of authority and justice in
    Kafka's 'In the Penal Colony' through a visceral and disturbing
    imagery.\closefatdquo
\end{description}
